---
title: "World wide natural disaster events — Analysis Report"
author: "Ziaul Hoque"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

### Approach

In this assignment, I selected a dataset from kaggle, regarding an analytical investigation of worldwide natural disaster occurrences documented between 2023 and 2025. This dataset contains important characteristics such the type of disaster, location, degree of severity, number of victims, financial losses, and emergency response timings are all captured here and also gives an opportunity to working with enormous numerical data.

My goal is to work with this data and get insights information through data driven approach and analyze incidents in different locations worldwide and get more information & visualize data to improve emergency response plans, risk mitigation techniques and disaster preparedness. So data driven insights are essential for identifying susceptible areas and allocating resources as emergency needed.

### Data Sources

<https://github.com/CandySoftInc/DATA607-Assignment1-Natural-Disaster-Events/blob/main/synthetic_disaster_events_2025.csv>

### Research Question

How we can improve more awareness to people before any natural disaster happens and how we can prevent loss or degree of severity, number of victims through this analysis?

### Codebase

**Including all related library**

```{r import relevant library}
library(tidyverse)
library(lubridate)
library(scales)
library(knitr)
```

**Import Dataset**

Loading dataset into dataframe to clean and preparing a dataset by modifying its columns into correct datatypes.

```{r loading dataset and preparing columns }
data_file <- "synthetic_disaster_events_2025.csv"
data <- read_csv(data_file, show_col_types = FALSE) |>
  mutate(
    date = as.Date(date),
    disaster_type = factor(disaster_type),
    location = factor(location),
    aid_provided = factor(aid_provided, levels = c("No", "Yes")),
    is_major_disaster = as.logical(is_major_disaster)
  )

cat("Rows:", nrow(data), "  Columns:", ncol(data))
```

Check Dataset Structure

```{r inspect data structure}

str(data)

head(data)
```

**Dataset Overview**

```{r dataset overview for events, economic loss, avg severity avg level & response time}
overview <- tibble(
  Measure = c(
    "Number of events",
    "Number of variables",
    "Earliest event date",
    "Latest event date",
    "Total affected population",
    "Total estimated economic loss",
    "Average severity",
    "Average response time (hours)"
  ),
  Value = c(
    nrow(data),
    ncol(data),
    format(min(data$date, na.rm = TRUE), "%Y-%m-%d"),
    format(max(data$date, na.rm = TRUE), "%Y-%m-%d"),
    comma(sum(data$affected_population, na.rm = TRUE)),
    dollar(sum(data$estimated_economic_loss_usd, na.rm = TRUE)),
    round(mean(data$severity_level, na.rm = TRUE), 2),
    round(mean(data$response_time_hours, na.rm = TRUE), 2)
  )
)
kable(overview, caption = "Dataset overview")
```

**Event Frequency**

```{r Disaster events by type}

events_by_type <- data %>% count(disaster_type, sort = TRUE)
ggplot(events_by_type, aes(y = reorder(disaster_type, n), x = n)) +
  geom_col() + coord_flip() +
  labs(y = NULL, x = "Number of events", title = "Disaster Events by Type")
```

```{r showing in table Disaster events by type}
knitr::kable(events_by_type, col.names = c("Disaster type", "Events"))
```

**Economic loss**

```{r economic loss mean, median value}
loss_by_type <- data %>%
  group_by(disaster_type) %>%
  summarise(
    events = n(),
    total_loss_usd = sum(estimated_economic_loss_usd, na.rm = TRUE),
    mean_loss_usd = mean(estimated_economic_loss_usd, na.rm = TRUE),
    median_loss_usd = median(estimated_economic_loss_usd, na.rm = TRUE),
    .groups = "drop"
  ) %>% arrange(desc(total_loss_usd))

ggplot(loss_by_type, aes(x = reorder(disaster_type, total_loss_usd), y = total_loss_usd)) +
  geom_col() + coord_flip() + scale_y_continuous(labels = dollar_format()) +
  labs(x = NULL, y = "Total estimated loss (USD)", title = "Estimated economic loss by disaster type")
```

```{r}
knitr::kable(loss_by_type, digits = 0)
```

**Major-disaster summary**

```{r amjor disaster summary}
major_summary <- data %>%
  group_by(is_major_disaster) %>%
  summarise(
    events = n(),
    avg_severity = mean(severity_level, na.rm = TRUE),
    avg_affected = mean(affected_population, na.rm = TRUE),
    avg_loss_usd = mean(estimated_economic_loss_usd, na.rm = TRUE),
    avg_response_hours = mean(response_time_hours, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(category = if_else(is_major_disaster == 1, "Major disaster", "Other disaster")) %>%
  select(category, everything(), -is_major_disaster)
knitr::kable(major_summary, digits = 2)
```

**Summary by location**

```{r total affected people, economic loss & avg severity by location}
location_summary <- data %>%
  group_by(location) %>%
  summarise(
    events = n(),
    total_affected = sum(affected_population, na.rm = TRUE),
    total_loss_usd = sum(estimated_economic_loss_usd, na.rm = TRUE),
    avg_severity = mean(severity_level, na.rm = TRUE),
    .groups = "drop"
  ) %>% arrange(desc(total_loss_usd))
knitr::kable(location_summary, digits = 2)
```

#### Outcome of Analysis from this report

1.  Which disaster types occur most frequently?
2.  How does the distribution of economic loss differ across disaster types?
3.  Is higher severity associated with greater estimated economic loss?
4.  How are response times related to the affected population?

#### My next Goal

Find relationships between among impact measures -

- Severity level vs. Economic loss and correlation coefficient value between these two factors.

- Response time vs. Affected population and correlation coefficient value between these two factors.
